\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-03 papers/jfp/3-abstractions.tex
Clarify the purpose of S3.8
Fix typos and formatting

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
Here is our first use of polymorpism in \hs{f}, discussed in \S\ref{sec-why-polymorphism}.
The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad and pass to the \hs{task}. The result has type \hs{Identity}~\hs{v}, which
we unwrap with \hs{runIdentity}.

@@

In this section we give examples of tasks whose definitions involve different
constraints on the computation context: \hs{Functor}, \hs{Applicative},
\hs{Monad} and \hs{MonadState}~\hs{s}.

>>>>

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
Here is our first use of polymorphism in \hs{f}, discussed in~\S\ref{sec-why-polymorphism}.
The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad and pass to the \hs{task}. The result has type \hs{Identity}~\hs{v}, which
we unwrap with \hs{runIdentity}.

@@

In this section we give examples of tasks whose definitions involve different
constraints on the computation context: \hs{Functor}, \hs{Applicative},
\hs{Monad} and \hs{MonadState}~\hs{s}. The purpose of these examples is to
continue building the intuition behind the \hs{Task} abstraction, and prepare
the reader for richer types of tasks that will appear
in~\S\ref{sec-implementations} and~\S\ref{sec-engineering}.

### 2019-05-03 papers/jfp/5-rebuilders.tex
Fix typos and formatting

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we pass the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

>>>>

\noindent
The function \hs{recordCT} is similar to \hs{recordVT}
from~\S\ref{sec-verifying-traces}, but instead of just passing
the hash of the resulting value, we pass the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

### 2019-05-03 papers/jfp/7-experience.tex
Fix typos and formatting

Based on the thinking in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies (\S\ref{sec-implementation-cloud}),
and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
In this section we tie these papers together, reflecting on our
experience of turning theory into practice.

@@

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using the data type:\footnote{
The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{1mm}
\noindent
In this paper we describe tasks more directly, in a monadic (or applicative or functorial)
style. But in fact the two are equivalent: \hs{Task Monad k v} is
isomporphic to \hs{Action k v v}.
To be concrete, the functions \hs{toAction} and \hs{fromAction} witness the
isomorphism in both directions.
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f

@@

\noindent for various choices of~\hs{c}, are studied in depth by~Jaskelioff and
O’Connor~\shortcite{jaskelioff2015representation}.

@@

\begin{itemize}
\item Two \hs{Ver} arguments are passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be specfied simultaneously, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

@@

\vspace{1mm}
\begin{center}
\parbox{0.8 \textwidth}{\emph{Thanks for the symlinks release, we just finished upgrading this build
system to use \cmd{-}\cmd{-share}. ... Building from scratch with a warm cache
takes around 5 seconds, saving us up to 2 hours. Not bad!}}
\end{center}
\vspace{1mm}

>>>>

Based on the theory developed in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
In this section we tie these papers together, reflecting on our
experience of turning theory into practice.

@@

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using the data type:\footnote{
The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{1mm}

\noindent
In this paper we describe tasks more directly, in a monadic (or applicative or functorial)
style. But in fact the two are equivalent: \hs{Task}~\hs{Monad}~\hs{k}~\hs{v} is
isomorphic to \hs{Action}~\hs{k}~\hs{v}~\hs{v}.
To be concrete, the functions \hs{toAction} and \hs{fromAction} defined below
witness the isomorphism in both directions.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f

@@

\noindent for various choices of~\hs{c}, are studied in depth by~Jaskelioff and
O'Connor~\shortcite{jaskelioff2015representation}.

@@

\begin{itemize}
\item Two \hs{Ver} arguments are passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be specified simultaneously, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

@@

\vspace{1mm}
\begin{center}
\parbox{0.8 \textwidth}{\emph{Thanks for the symlinks release, we just finished
upgrading this build system to use \textsf{-}\textsf{-share}. ... Building from
scratch with a warm cache takes around 5 seconds, saving us up to 2 hours. Not
bad!}}
\end{center}
\vspace{1mm}

### 2019-05-03 papers/jfp/8-engineering.tex
Fix typos and formatting

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Fetch k f = @\std{forall}@ v. k v -> f v

newtype TaskT c k v = TaskT { runT :: @\std{forall}@ f. c f => Fetch k f -> f v }

@@

Some build tasks produce multiple output keys -- for example \cmd{ghc A.hs} produces both
\cmd{A.hi} and \cmd{A.o}. This pattern can be encoded by having a key \cmd{A.hi+A.o} which produces
both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety from
\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:

@@

task :: TasksT Monad Key
task (File "A.hi") = Just $ TaskT $ \fetch -> do
    [hi, o] <- fetch (Files ["A.hi","A.o"])
    pure hi
task (File "A.o") = Just $ TaskT $ \fetch -> do
    [hi, o] <- fetch (Files ["A.hi","A.o"])
    pure o
task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
    -- run ghc here
\end{minted}
\vspace{1mm}

The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for exactly the same
\cmd{Files} key. If one \hs{File} key swapped the order of the list to \hs{Files} then it would likely
run \cmd{ghc} twice. To help construct a well-formed multiple-output build task
it is convenient to partition the set of keys into \emph{elementary subsets}.
We can encode such a partition with a function from any key to all the members of its
subset (in a consistent order).

@@

\noindent In our example the partition function on either of the output names \hs{"A.hi"} or \hs{"A.o"} would
return \hs{["A.hi", "A.o"]}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
task :: Partition FilePath -> TasksT Monad Key
task partition (File k) = Just $ TaskT $ \fetch -> do
    let ks = partition k
    let Just i = elemIndex k ks
    vs <- fetch (Files ks)
    pure (vs !! i)
task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
    -- run ghc here
-- more rules on elementary subsets as required
\end{minted}
\vspace{1mm}

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{text}
A1 = 20      B1 = A1 + A2
A2 = 10
\end{minted}
\vspace{0.5mm}

>>>>

\vspace{1mm}
\begin{minted}[xleftmargin=5pt]{haskell}
type Fetch k f = @\std{forall}@ v. k v -> f v

newtype TaskT c k v = TaskT@\,@{@\,@runT :: @\std{forall}@ f. c f => Fetch k f -> f v@\,@}

@@

Some build tasks produce multiple output keys -- for example \cmd{ghc A.hs} produces both
\cmd{A.hi} and \cmd{A.o}. This pattern can be encoded by having a key \cmd{A.hi+A.o} which produces
both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety
from~\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:

@@

task :: TasksT Monad Key
task (File "A.hi") = Just $ TaskT $ \fetch -> do
    [hi,o] <- fetch (Files ["A.hi","A.o"])
    return hi
task (File "A.o") = Just $ TaskT $ \fetch -> do
    [hi,o] <- fetch (Files ["A.hi","A.o"])
    return o
task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
    compileWithGHC "A.hs"
\end{minted}
\vspace{1mm}

The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for
exactly the same \hs{Files} key. If one \hs{File} key swapped the order of the
list to \hs{Files} then it would likely run GHC twice. To help construct a
well-formed multiple-output build task it is convenient to partition the set of
keys into \emph{elementary subsets}. We can encode such a partition with a
function from any key to all the members of its subset (in a consistent order).

@@

\noindent
In our example the partition function on either of the output names \hs{"A.hi"}
or \hs{"A.o"} would return \hs{["A.hi","A.o"]}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
task :: Partition FilePath -> TasksT Monad Key
task partition (File k) = Just $ TaskT $ \fetch -> do
    let ks = partition k
    let Just i = elemIndex k ks
    vs <- fetch (Files ks)
    return (vs !! i)
task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
    compileWithGHC "A.hs"
... -- more tasks for elementary subsets as required
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1 = 20      B1 = A1 + A2
A2 = 10
\end{minted}
\vspace{1mm}
